---
title: "Week 3 In-Class Lab"
author: "Tiffany Tang"
date: "`r format(Sys.time(), '%B %d, %Y')`"
header-includes:
   - \usepackage{float}
output: 
  html_document:
    number_sections: true
---

# Goals

In this in-class lab, we will 

- play around with different  ```ggplot``` themes and color schemes
- learn about a few useful visualization packages that have different capabilities from ```ggplot```

```{r setup, echo = FALSE, warning = FALSE, message = FALSE}
# setting default knitr options
knitr::opts_chunk$set(
  echo = FALSE,  
  warning = FALSE,  
  message = FALSE,  
  fig.out = "100%",
  fig.align = "center", 
  fig.pos = "H",
  cache = FALSE)

# load useful libraries
library(tidyverse)
library(R.utils)
library(GGally)
library(superheat)  # Note: installed from GitHub, not CRAN
library(viridis)
library(RColorBrewer)
library(ggridges)
library(scales)
library(gridExtra)
library(ggpubr)
```

```{r source-scripts}
# source all files in the R/ directory
sourceDirectory("./R/", modifiedOnly = F, recursive = F) # useful functions
```

To guide you through this material, I will provide examples using the ```gapminder``` data that we worked with last week. Then you will work through similar problems using the ```perceptions``` data. Put briefly, the perceptions data deals with the perceptions of different words relating to probabilities and numbers. The raw data came from /r/samplesize responses to the following question: What [probability/number] would you assign to the phrase "[phrase]"? You can read more about the perceptions data at (https://github.com/zonination/perceptions).

# Themes and Color Schemes

As I mentioned in the slides, it can become monotonous to look at 100+ plots with the same gridded gray ```ggplot``` background and the same default ```ggplot``` color scheme. A simple way to mix things up is to apply a different built-in ```ggplot``` theme. Pick your favorite \href{https://ggplot2.tidyverse.org/reference/ggtheme.html}{here} or simply google "custom ggplot themes" for a plethora of options.

```{r load-data}
# load the original gapminder data
gapminder_orig <- loadGapminderData()
# clean the gapminder data
gapminder <- cleanGapminderData(gapminder_orig)

# load the original perceptions data
probly_orig <- loadPerceptData(filename = "probly.csv")  # prob survey
numberly_orig <- loadPerceptData(filename = "numberly.csv")  # number survey
# clean the perceptions data
probly <- cleanProblyData(probly_orig)
numberly <- cleanNumberlyData(numberly_orig)
```

You can play around with the built-in ggplot themes or even build a custom theme if you're feeling ambitious.

```{r themes}

# default plot
base_plt <- gapminder %>%
  # remove Oceania
  filter(continent != "Oceania") %>%
  # input data frame into ggplot base function
  ggplot() +
  # make a separate plot for each continent
  facet_wrap(~continent) +
  # add a boxplot layer grouped by year
  geom_boxplot(aes(x = year, y = life_exp, group = year)) +
  xlab("Year") +
  ylab("Life expectancy") +
  ggtitle("Life expectancy over time")

# view default theme
base_plt +
  ggtitle("Default Theme")

# add built-in theme
base_plt + 
  theme_classic() +
  ggtitle("Classic Built-in Theme")

# add custom theme (see utils.R for definition of z_theme())
base_plt +
  z_theme() +
  ggtitle("Custom Theme")

```

1. Compute the mean probability associated with each phrase in the ```probly``` data and plot the results in a bar graph. Add a built-in or custom ggplot theme to this plot. In your plot, make sure you can read the x-axis/phrase labels.

```{r theme-probly}
my_theme <- theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

probly %>%
  group_by(phrase) %>%
  summarise(mean.prob = mean(prob)) %>%
  ggplot() +
  aes(x = phrase, y = mean.prob) +
  geom_bar(position = "dodge", stat = "identity") +
  labs(x = "Phrase", y = "Mean Probability") +
  my_theme

```



# Color Schemes

Choosing an appropriate color scheme for your plots can drastically improve the readability of your plots. Sometimes, it is worthwhile to stray from the default colors in ggplot. The ```viridis``` package, in particular, has several nice continuous color schemes and can be easily applied to ```ggplot``` objects (see ```scale_color_virids()``` or ```scale_fill_viridis()```). Aside from ```viridis```, the package ```RColorBrewer``` has some nice color palettes, and if you're incredibly ambitious, you can even create your own color palettes. 


```{r color-schemes-cts}

base_plt <- ggplot(gapminder) +
  geom_point(aes(x = log10(life_exp), y = log10(gdp_per_cap), 
                 color = year),
             size = .5, alpha = .8)

# base plot
base_plt +
  ggtitle("Default Color Scheme (Continuous)")

# viridis
base_plt + 
  scale_color_viridis_c(option = "C") +  # try changing the option argument
  ggtitle("Viridis Color Scheme (Continuous)")

```


```{r color-scheme-discrete}

base_plt <- gapminder %>%
  filter(continent != "Oceania") %>%
  ggplot() +
  geom_point(aes(x = log10(life_exp), y = log10(gdp_per_cap), 
                 color = continent),
             size = .5, alpha = .8)

# base plot
base_plt +
  ggtitle("Default Color Scheme (Discrete)")

# RColorBrewer
my_palette <- brewer.pal(n = nlevels(gapminder$continent), 
                         name = "Dark2")  # try different palettes
base_plt + 
  scale_color_manual(values = my_palette) +
  ggtitle("RColorBrewer Color Scheme (Discrete)")

```


2. Using the ```numberly``` data, make a boxplot to visualize the distribution of assigned numbers corresponding to each phrase (i.e., there should be a boxplot for each phrase). Due to the highly skewed distributions, please take the $\log_{10}$-transform of the numeric data. Then, add an additional layer to the plot using ```geom_jitter()``` to explicitly plot the assigned numbers for each phrase. Finally, add an appropriate (non-default) color scheme to your plot.

```{r color-numberly}

ggplot(numberly, aes(phrase, number)) +
  geom_boxplot(aes(fill = phrase), alpha = 0.5) +
  geom_jitter(aes(color = phrase), size = 1, alpha = .45) +
  scale_y_log10(labels = trans_format("log10", math_format(10^.x)),
                breaks = 10^(-2:6)) +
  guides(fill = FALSE, color = FALSE) +
  scale_color_viridis_d(option = "C", begin = 0, end = .9) +
  scale_fill_viridis_d(option = "C", begin = 0, end = .9) +
  labs(title = "Perceptions of Probability",
       x = "Phrase",
       y = "Assigned Number",
       caption = "created by /u/zonination") +
  coord_flip() +
  z_theme()

```


# Heatmaps

Rather than scatterplots, another type of graph that can often be very informative is a heatmap. Rebecca Barter, a student of Bin and a former 215A GSI, developed the ```superheat``` package, which can be used to make nice heatmaps. In Figure \ref{fig:heatmap}, we plot a heatmap of the life expectancy across time for various countries. Note that the countries have been clustered via hierarchical clustering with Ward's linkage (don't worry if you have no idea what this means yet). This allows us to more easily see patterns in the life expectancies across different groups of countries. Note the difference between the clustered heatmap and the heatmap without clustering. In the clustered heatmap, some of the clusters correspond to coherent geographical regions, which makes intuitive sense.

```{r heatmaps, fig.height = 8}

# convert life expectancy data to wide format for superheat
life_exp_wide <- gapminder %>%
  select(country, year, life_exp) %>%
  spread(key = "year", value = "life_exp") %>%
  sample_frac(size = .5)  %>% # optional; for readability/illustration only
  arrange(country) %>%
  column_to_rownames("country")

# ordinary heatmap 
superheat(life_exp_wide,
          pretty.order.rows = FALSE,
          clustering.method = "hierarchical",
          linkage.method = "ward.D",
          force.left.label = TRUE,
          left.label.text.size = 2.5,
          bottom.label.text.size = 2.5, 
          title = "Life Expectancy (in years)")

# heatmap with clustered countries and mean life expectancy on top
superheat(life_exp_wide, 
          # cluster the countries
          pretty.order.rows = TRUE,
          row.dendrogram = T,
          clustering.method = "hierarchical",
          linkage.method = "ward.D",
          # plot mean life expectancy per year on top
          yt = colMeans(life_exp_wide),
          yt.plot.size = .2,
          yt.axis.name = "Mean Life \n Expectancy",
          yt.axis.name.size = 8,
          # add pretty labels
          force.left.label = TRUE,
          left.label.text.size = 2.5,
          bottom.label.text.size = 2.5, 
          title = "Life Expectancy (in years)")

```

3. Use ```superheat``` to create a heatmap for the ```probly``` data, where the samples are on the x-axis, phrases on the y-axis, and probabilities represented by the heatmap values/colors. Cluster the samples using hierarchical clustering with complete linkage but don't plot the resulting dendrogram. To the right of the heatmap, plot a bar graph with the average probabilities for each phrase.

```{r superheat-probly}

# convert long to wide
probly_wide <- probly %>%
  spread(key = "phrase", value = "prob") %>%
  column_to_rownames("id")

superheat(as.data.frame(t(probly_wide)), 
          # cluster the samples
          pretty.order.cols = TRUE,
          clustering.method = "hierarchical",
          linkage.method = "complete",
          # plot mean probabilities on the right
          yr = colMeans(probly_wide),
          yr.plot.type = "bar",
          yr.axis.name = "Mean Prob.",
          yr.axis.name.size = 8,
          # add pretty labels
          left.label.text.size = 2.5,
          bottom.label = "none",
          title = "Perception of Probabilities")
```

# Pair Plots

Last week, we looked at the relationship between life expectancy over time and gdp vs. life expectancy, but we did so separately. It may be informative to look at multiple pair-wise relationships in the data in a single plot. In the ```GGally``` package, the ```ggpairs()``` function allows us to plot a matrix of pair plots, showcasing many different pair-wise relationships in the dataset at the same time.

```{r pair-plots}

# a very basic pair plot
ggpairs(gapminder, 
        columns = c(3, 5, 6),
        title = "A Very Basic Pair Plot")

# show pairwise relationships among the continuous features in data
ggpairs(data = gapminder %>% mutate(population = log10(population),
                                    gdp_per_cap = log10(gdp_per_cap)),
        # which columns to plot
        columns = c(3, 5, 6),
        # set color aesthetic
        mapping = aes(color = continent),
        # set geom_point(alpha = , size = ) for continuous vars in lower plots
        lower = list(continuous = wrap("points", alpha = .5, size = .25)),  
        # set geom_density(alpha = ) for continuous vars in diagonal plots
        diag = list(continuous = wrap("densityDiag", alpha = .5)),
        # make correlation text smaller in upper diag plots
        upper = list(continuous = wrap("cor", size = 3)),
        title = "A Fancier Pair Plot")

```


4. Going beyond pair-wise compairsons between continuous variables, ```ggpairs``` also allows for discrete variables in the pair plots. Create the same type of pair plot as above, but include the following four variables from ```gapminder```: population, continent, life expectancy, and GDP per capita. Also, add a theme to your pair plot.

```{r pair-plot-discrete}

# can also see relationships among continuous and discrete/categorical features
ggpairs(data = gapminder %>% mutate(population = log10(population),
                                    gdp_per_cap = log10(gdp_per_cap)),
        columns = 3:6,
        mapping = aes(color = continent),
        # set geom_point(alpha = , size = ) for continuous vars in lower plots
        lower = list(continuous = wrap("points", alpha = .5, size = .15)),  
        # set geom_density(alpha = ) for continuous vars in diagonal plots
        diag = list(continuous = wrap("densityDiag", alpha = .5)),
        # make correlation text smaller in upper diag plots
        upper = list(continuous = wrap("cor", size = 2)),
        title = "An Even Fancier Pair Plot") + 
  theme_light()

```


There are a lot of additional options that you can set in the ```ggpairs()``` function, so check out the help page for ```ggpairs```. This help page is very informative, and you can do a lot with the ```ggpairs``` function.


# Ridgeline Plots/Joyplots

The ```ggridges``` package provides an additional ```geom``` attribute called ```geom_density_ridges``` that can be added to ```ggplot``` objects. ```geom_density_ridges``` arranges multiple density plots in a staggered fashion.

```{r ggridges}

ggplot(gapminder) +
  aes(x = log(population), y = continent, fill = continent) +
  geom_density_ridges(alpha = .5) +
  z_theme()

```

5. Use ```geom_density_ridges``` to make a nice visualization of the ```numberly``` data. Make sure to take care of skewness and add an appropriate (non-default) color scheme. Please also remove the legend.

```{r ggridges-numberly}

num_densities <- ggplot(numberly) +
  aes(x = log10(number), y = phrase, fill = phrase) +
  geom_density_ridges(alpha = .5, size = .3) +
  scale_fill_viridis_d(option = "C") +
  labs(x = "log10(Assigned Number)", y = "Phrase", 
       fill = "Phrase", title = "Perception of Numbers") +
  guides(fill = F) +
  theme_minimal() +
  theme(axis.title = element_text(face = "bold"),
        legend.title = element_text(face = "bold"),
        title = element_text(face = "bold"))

num_densities

```


# Side-by-side Plots

Sometimes, it may be useful to organize multiple plots side-by-side. Two packages for doing so are ```gridExtra::grid.arrange``` and ```ggpubr::ggarrange```.

6. One of the advantages of using ```ggarrange``` over ```grid.arrange``` is the ease of creating a common legend and subplot labels with ```ggarrange```. Below, create two scatterplots using the ```gapminder``` data: the first scatterplot showing population vs. life expectancy and the second scatterplot showing gdp per capita vs life expectancy. Color the points in both scatterplots by continent and place these scatterplots side-by-side using ```ggarrange```. Since both scatterplots have the same color legend, use a common legend and set the legend's position to "below". Also, set the ```labels``` argument to "AUTO".

```{r ggarrange, out.width = "100%", results = "hide"}

plt1 <- ggplot(gapminder) +
  aes(x = log10(population), y = life_exp, color = continent) +
  geom_point(size = .5, alpha = .7) +
  labs(title = "Population vs Life Expectancy", color = "Continent",
       x = "log(Population)", y = "Life Expectancy (yrs)") +
  theme_light()
plt2 <- ggplot(gapminder) +
  aes(x = log10(gdp_per_cap), y = life_exp, color = continent) +
  geom_point(size = .5, alpha = .7) +
  labs(title = "GDP Per Capita vs Life Expectancy", color = "Continent",
       x = "log(GDP Per Capita)", y = "Life Expectancy (yrs)") +
  theme_light()
ggarrange(plt1, plt2, ncol = 2, nrow = 1,
          labels = "AUTO", legend = "bottom", common.legend = T)

```


7. Unlike ```ggarrange```, ```grid.arrange``` allows for very flexible plotting layouts. In addition to the two subplots that you created in exercise 6, create a third plot using ```geom_bar``` which shows the number of data points (i.e., countries) from each continent. Use ```grid.arrange``` to create a plot with the population vs life expectancy plot in the upper left quadrant, the GDP per capita vs life expectancy plot in the upper right quadrant, and the bar plot in the lower two quadrants. The end result should be similar to the plot shown in the gridExtra::grid.arrange slide from class.

```{r gridExtra, out.width = "100%", fig.height = 5}

plt3 <- ggplot(gapminder) +
  aes(x = continent, fill = continent) +
  geom_bar() +
  guides(fill = F) +
  labs(x = "Continent", y = "Frequency") +
  theme_light()
grid.arrange(plt1, plt2, plt3, 
             layout_matrix = matrix(c(1, 2, 3, 3), nrow = 2, byrow = T))

```


# Interactive Plots

Challenge Exercise: Come up with your own interactive visualization of the perception data. Be creative!


